Replicator-mutator and noise-perturbed replicator dynamics

### 2014-09-21 snippets/data_exploration/data_exploration_tmp.Rmd
data exploration with varying tolerance; some plotting functions added

require('ellipse')
require('ggplot2')
require('reshape2')
source('~/Desktop/data/svn/programming/R/helpers/helpers.R')

data = read.csv('~/Desktop/data/svn/vagueness-games/results/20140918-142408.csv', 
                colClasses=c('numeric','factor','numeric','numeric','numeric',
                             'factor','numeric','numeric','numeric','numeric',
                             'numeric','numeric','numeric','numeric','numeric',
                             'numeric','numeric','numeric','character','character'))

@@

attach(ds)
```

@@

pairs(~ Number.of.states + Impairment + Tolerance + Speaker.entropy + Hearer.entropy + Speaker.convexity +
        Speaker.Voronoiness + Hearer.Voronoiness +
        Speaker.informativity + Hearer.informativity + Expected.utility + Iterations +
        Speaker.Convex.Cat, data=ds)
```

@@

dms = summarySE(dm, groupvars = c("Number.of.states", "Impairment", "Tolerance", "variable"), measurevar="value")
dsub = subset(dms, !grepl("informativity",dms$variable) )

@@

>>>>

require('ellipse')
require('ggplot2')
require('reshape2')
require('plyr')
source('~/Desktop/data/svn/programming/R/helpers/helpers.R')
source('~/Desktop/data/svn/vagueness-games/data-analysis.R')

data = read.csv('~/Desktop/data/svn/vagueness-games/results/20140918-142408.varTolerance.csv', 
                colClasses=c('numeric','factor','numeric','numeric','numeric',
                             'factor','numeric','numeric','numeric','numeric',
                             'numeric','numeric','numeric','numeric','numeric',
                             'numeric','numeric','numeric','character','character'))

@@

attach(ds)

plot.language.byindex = function(dataFrame,rowindex){
  plot.language(dataFrame[rowindex,]$Speaker.strategy.file,
                dataFrame[rowindex,]$Hearer.strategy.file)
}

plot.language.MinMaxProperty = function(dataFrame,colindex){
  show(names(dataFrame)[colindex])
  show(summary(dataFrame[,colindex]))
  p = arrangeGrob(plot.language.byindex(dataFrame,which.min(dataFrame[,colindex])),
                  plot.language.byindex(dataFrame,which.max(dataFrame[,colindex])))
}

@@

pairs(~ Number.of.states + Impairment + Tolerance + Speaker.entropy + Hearer.entropy + 
        Speaker.Voronoiness + Hearer.Voronoiness +
        Speaker.informativity + Hearer.informativity + Expected.utility + Iterations +
        Speaker.Convex.Cat, data=ds)
```

@@

dms = summarySE(dm, groupvars = c("Number.of.states", "Impairment", "Tolerance", "variable"), measurevar="value")
dsub = dms

@@

Informativity
-------------

This is the notion from Skyrms. It seems very well-behaved. There is a clear positive correlations between speaker and hearer informativity.

```{r}
cor.test(data$Speaker.informativity, data$Hearer.informativity)
qplot(data$Speaker.informativity, data$Hearer.informativity) + geom_smooth(method=lm)
```

There is a clear negative correlation between impairment and informativity. The remaining variance here must and can be accounted for in terms of Number.of.states (see main plot).

```{r}
cor.test(data$Impairment, data$Speaker.informativity)
qplot(data$Impairment, data$Speaker.informativity) + geom_smooth(method=lm)
```

Let's look at the 50 state case, with impairment 0.1 and Tolerance 0.2.

```{r}
d = subset(data, Impairment == 0.05 & Number.of.states == 50 & Tolerance == 0.2)
colindex = which(names(d) == "Speaker.informativity")
show(plot.language.MinMaxProperty(d,colindex))
```

How much identity in outcomes?
------------------------------

I'm curious how much different properties actually differ within each triplet of dependent variables (tolerance, impairment, number of states). We can plot the difference between the maximmal and the minimal value of each property for each triplet of independent variables.

dm = melt(subset(ds,Iterations < 205)[,-which(names(ds) == "Iterations")],
          id.vars = c("Number.of.states", "Impairment", "Tolerance")) # melted data

MinMaxProperty = ddply(dm, .(Number.of.states, Impairment, Tolerance, variable), 
                      summarise, 
                      MaxMinDiff = max(value) - min(value),
                      MinProperty = min(value), MaxProperty = max(value), MeanProperty = mean(value))

pd <- position_dodge(.01)
sp = ggplot(MinMaxProperty, aes(x=Impairment, y=MaxMinDiff, colour=factor(Number.of.states))) + 
     geom_point(position = pd)

sp + facet_grid(variable ~ Tolerance, scales="free")
```

We can then look at the maximal difference among all properties. (Removing Voronoiness here, because it screws up the y-axis dimensions.)

```{r}
diffTrials = ddply(subset(MinMaxProperty, variable != "Speaker.Voronoiness" & variable != "Hearer.Voronoiness"),
                   .(Number.of.states, Impairment, Tolerance), 
                  summarise, 
                  diffTrials = max(MaxMinDiff) <=1e-5,
                  maxDiffTrials = max(MaxMinDiff) )

ggplot(diffTrials, aes(x=Impairment, y=maxDiffTrials, )) + 
     geom_point() + facet_grid(Number.of.states ~ Tolerance, scales="free")

We see clearly that, all else equal (e.g. fixed tolerance and number of states), impairment 'unifies' evolutionary outcomes. The outcomes are all almost identical (in terms of the recorded properties).

But we should check whether this 'uniformity' is different from just convexity vs. non-convexity.

d = subset(dsub, dsub$variable =="Speaker.Convex.Cat")

pd <- position_dodge(.01)
sp = ggplot(d, aes(x=Impairment, y=1-value )) + 
     geom_point(position = pd)

sp + facet_grid(Number.of.states ~ Tolerance, scales="free")

By visual inspection, there are differences in evolutionary outcomes that not due to convexity issues.
